Learning Discrete-valued Bayesian Networks from Mixed Data

### 2015-11-18 BN_discretizer_paper.tex
Rewrite graph1, center plots, add content to experiment part

\begin{figure}[ht]
  \label{fig:example_factorization}
  \begin{tabular}{cc}
    \input{graph1}
  \end{tabular}
  \caption{Factorization of $P(D \mid M)$}
\end{figure}

@@

\begin{algorithm}
\caption{Discretization of one continuous variable in a Bayesian network}
\label{alg:disc_one}
\begin{algorithmic}[5]
\Function{Discretize}{$D$, $G$}
\State
\State $N \leftarrow$ the number of instances
\State $N' \leftarrow$ the number of unique values in $D_X$
\State $H \leftarrow$ an $N \times N$ matrix such that $H[u,v] = h(u,v)$ per Algorithm~\ref{alg:h} in the appendix
\State $b \leftarrow$ the increasing sequence of $M$ integers as previously defined
\State $L \leftarrow$ the largest cardinality over all discrete variables in the Markov blanket of $X$
\State $S[u] \leftarrow$ the optimal objective value of subproblems with instances from $1$ to $u$
\State $M[u] \leftarrow$ the optimal discretization for subproblems with instances from $1$ to $u$
\State $W[u]  \leftarrow - \log\brock{1 - {\exp\paren{- L \cdot{ {{x_{b(u)+1} - x_{b(u)}}\over{x_N - x_1}}}}}}$ for $u = 1,2, \ldots,N'-1$ and $L(N') \leftarrow 0$
\State
\For {$v = 1$ to $N'$}
\If {$v = 1$}
\State $S[v] \leftarrow g \left(1,b[v] \right) + L[v]$
\State $M[v] \leftarrow \{ ({x_{b[v]} + x_{b[v]+1}}) / 2\}$
\Else
\State $s \leftarrow \infty$ and $boundary \leftarrow \infty$
\For {$u = 1$ to $v$}
\State $s' \leftarrow S[u] + g \left( b[u]+1,b[v] \right) +  {L \cdot {{x_{b[v]} - x_{b[u] + 1}}\over{x_N - x_1}}} + W[v]$
\If {$s' < s$}
\State $s \leftarrow s'$
\State $boundary \leftarrow ({x_{b[u]} + x_{b[u]+1}}) / 2$
\EndIf
\EndFor
\State $S[v] \leftarrow s$
\State $M[v] \leftarrow M[u] \cup \{ boundary\}$
\EndIf
\EndFor
 \State \Return $M$
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

\begin{equation}
\begin{aligned}
\frac{1}{2} \log(N) \left\lbrace  \lvert \Pi_{X_i} \rvert (\lvert X^*_i \rvert - 1) +
 {\sum_{j,X_i \in \Pi_{X_i}}} \lvert \Pi_{X^*_j} \rvert (\lvert X_j \rvert - 1) \right\rbrace + \log(\lvert {X^*_i} \rvert) \\
 + (N_i-1) H \left( {{|X^*_i| - 1}\over{N_i -1}}  \right) -N \cdot \left[ I(X^*_i,\Pi_{X_i}) + {\sum_{j,X_i \in \Pi_{X_i}}} I(X_j, \Pi_{X^*_j}) \right]\text{,}
\end{aligned}
\end{equation}

@@

\begin{figure}[ht]
      \input{graph_auto_exp1_distr_1_3}
  \caption{Comparison of the raw data and discretization result on variable $A$ and $C$ by the optimal Bayesian approach. The marginal probabilities of discretized areas match the distribution of the taw data very well.}
  \label{fig:auto_exp1_distr_1_3}
\end{figure}

@@

\begin{figure}[ht]
      \input{graph_auto_exp2_distr_1_3}
  \caption{Comparison of the raw data and discretization result on variable $A$ and $C$ in Figure \ref{fig:auto_graph_2}. Although the number of discretization edges is less than in Figure \ref{fig:auto_exp1_distr_1_3}, it still matches the raw data distribution very well.}
  \label{fig:auto_exp2_distr_1_3}
\end{figure}

@@

\begin{figure}[ht]
  \centering
      \input{graph_wine_1}
  \caption{Structure from prediscretized K2 structure learning to compare the methods for Wine dataset}
  \label{fig:wine_graph_1}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \begin{tabular}{cc}
   \input{graph_housing_1}
   \end{tabular}
  \caption{Structure from prediscretized K2 structure learning to compare the methods for Housing dataset}
  \label{fig:housing_graph_1}
\end{figure}

\begin{table}
  \centering\caption{Discretization result of Housing dataset based on the graph in Figure~\ref{fig:housing_graph_1}. The first twelve rows show the numbers of intervals after discretization by the three method. The thirteenth row is the log-likelihood of the dataset based on the discretizatio results. The heuristic method is over sensitive and discretizes some variables into too many intervals. The MDL method is less sensitive and discretizes most variables by only one interval. The optimal Bayesian approach has a good control on number of intervals. \tim{Fig. vs. Figure}}
  \scalebox{1.0}{
  \input{table_housing_1.tex}
  }

@@

\begin{figure}[ht]
    % \begin{tabular}{cc}
      \input{graph_housing_exp1_distr_3_5}
    % \end{tabular}
  \caption{The discretization result on variable $C$ and $E$ of the network in Figure~\ref{fig:housing_graph_1}. The scatter points are jittered due to the repeated values in the dataset. Furthermore, these repeated values lead to high number of discretization edges, since repeated values can easily form discretized areas by their own.}
  \label{fig:housing_exp1_distr_3_5}
\end{figure}

\begin{figure}[ht]
     %\begin{tabular}{cc}
    \input{graph_housing_exp1_distr_8_5}
     %\end{tabular}
  \caption{The discretization result on variable $H$ and $E$ of the network in Figure~\ref{fig:housing_graph_1}. Data points are not over repeated on $H$, thus the number of edges on $H$ remains desirable.}
  \label{fig:housing_exp1_distr_8_5}
\end{figure}

@@

\begin{figure}[ht]
    % \begin{tabular}{cc}
      \input{graph_housing_2}
    % \end{tabular}
  \caption{The learned discrete-valued Bayesian network from the Housing dataset. Variables $C$, $E$, $J$, $K$ have high numbers of intervals after discretization, since they are in a Markov blanket.
  \tim{Var with 1 interval?}
  }
  \label{fig:housing_graph_2}
\end{figure}

@@

\begin{figure}[ht]
     %\begin{tabular}{cc}
      \input{graph_housing_exp2_distr_3_5}
     %\end{tabular}
  \caption{The discretization result on variable $C$ and $E$ of the network in Figure~\ref{fig:housing_graph_2}. The high number of discretization edges is due to the repeated values on the dataset in variable $C$ and $E$.}
  \label{fig:housing_exp2_distr_3_5}
\end{figure}

@@

To quantatively discuss the sensitivity of the three methods, consider the Bayesian network in Figure \ref{fig:exp_discuss}, where $X$ is continuous-valued and $P_1$ and $P_2$ is discrete-valued. Let $P = \{  P_1, P_2 \}$. Assume $D_X = \{1.0,2.0,3.0, \ldots N \}$. Therefore, the corresponding objective functions to discretize $X$ in the three methods are:
\begin{equation}
\begin{aligned}
\text{Obj}_{\text{MDL}} & = Z_1 \cdot k + \log(N) \cdot \sum^k_{i=1} \sum^{J_P}_{j=1} I(a^*_i, (p_1,p_2)_j) \\
\text{Obj}_{\text{Bayesian}} &= Z_2 \cdot k + \log {{n_i + J_P - 1}\choose{J_P -1}} + \sum^k_{i=1} \log \left(   {{n_i !}\over{n^{(P)}_{i,1}! n^{(P)}_{i,2}! \cdots n^{(P)}_{i,J_P}!    }} \right) \\
\text{Obj}_{\text{Heuristic}} &= Z_2 \cdot k + \log {{n_i + J_{P_1} - 1}\choose{J_{P_1} -1}} + \log {{n_i + J_{P_2} - 1}\choose{J_{P_2} -1}} +\\
& \sum^k_{i=1}  \left[ \log \left(   {{n_i !}\over{n^{(P_1)}_{i,1}! n^{(P_1)}_{i,2}! \cdots n^{(P_1)}_{i,J_{P_1}}    }} \right)  + \log \left(   {{n_i !}\over{n^{(P_2)}_{i,1}! n^{(P_2)}_{i,2} \cdots n^{(P_2)}_{i,J_{P_2}} }} \right) \right]
\end{aligned}
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{acknowledgements}
%If you'd like to thank anyone, place your comments here
%and remove the percent signs.
%\end{acknowledgements}

>>>>

\begin{figure}[ht]
\centering
  \label{fig:example_factorization}
  %\begin{tabular}{cc}
    \input{graph1}
  %\end{tabular}
  \caption{Factorization of $P(D \mid M)$}
\end{figure}

@@

\begin{algorithm}
\caption{Discretization of one continuous variable in a Bayesian network}
\label{alg:disc_one}
\begin{algorithmic}[5]
\Function{Discretize}{$D$, $G$}
\State $N \leftarrow$ the number of instances
\State $N' \leftarrow$ the number of unique values in $D_X$
\State $H \leftarrow$ an $N \times N$ matrix such that $H[u,v] = h(u,v)$ per Algorithm~\ref{alg:h} in the appendix
\State $b \leftarrow$ the increasing sequence of $M$ integers as previously defined
\State $L \leftarrow$ the largest cardinality over all discrete variables in the Markov blanket of $X$
\State $S[u] \leftarrow$ the optimal objective value of subproblems with instances from $1$ to $u$
\State $M[u] \leftarrow$ the optimal discretization for subproblems with instances from $1$ to $u$
\State $W[u]  \leftarrow - \log\brock{1 - {\exp\paren{- L \cdot{ {{x_{b(u)+1} - x_{b(u)}}\over{x_N - x_1}}}}}}$ for $u = 1,2, \ldots,N'-1$ and $L(N') \leftarrow 0$
\For {$v \leftarrow 1$ to $N'$}
\If {$v = 1$}
\State $S[v] \leftarrow g \left(1,b[v] \right) + L[v]$
\State $M[v] \leftarrow \{ ({x_{b[v]} + x_{b[v]+1}}) / 2\}$
\Else
\State $s \leftarrow \infty$ and $boundary \leftarrow \infty$
\For {$u \leftarrow 1$ to $v$}
\State $s' \leftarrow S[u] + g \left( b[u]+1,b[v] \right) +  {L \cdot {{x_{b[v]} - x_{b[u] + 1}}\over{x_N - x_1}}} + W[v]$
\If {$s' < s$}
\State $s \leftarrow s'$
\State $boundary \leftarrow ({x_{b[u]} + x_{b[u]+1}}) / 2$
\EndIf
\EndFor
\State $S[v] \leftarrow s$
\State $M[v] \leftarrow M[u] \cup \{ boundary\}$
\EndIf
\EndFor
 \State \Return $M$
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

\begin{equation}
\label{eqn:MDL}
\begin{aligned}
\frac{1}{2} \log(N) \left\lbrace  \lvert \Pi_{X_i} \rvert (\lvert X^*_i \rvert - 1) +
 {\sum_{j,X_i \in \Pi_{X_i}}} \lvert \Pi_{X^*_j} \rvert (\lvert X_j \rvert - 1) \right\rbrace + \log(\lvert {X^*_i} \rvert) \\
 + (N_i-1) H \left( {{|X^*_i| - 1}\over{N_i -1}}  \right) -N \cdot \left[ I(X^*_i,\Pi_{X_i}) + {\sum_{j,X_i \in \Pi_{X_i}}} I(X_j, \Pi_{X^*_j}) \right]\text{,}
\end{aligned}
\end{equation}

@@

\begin{figure}[ht]
\centering
      \input{graph_auto_exp1_distr_1_3}
  \caption{Comparison of the raw data and discretization result on variable $A$ and $C$ by the optimal Bayesian approach. The marginal probabilities of discretized areas match the distribution of the taw data very well.}
  \label{fig:auto_exp1_distr_1_3}
\end{figure}

@@

\begin{figure}[ht]
\centering
      \input{graph_auto_exp2_distr_1_3}
  \caption{Comparison of the raw data and discretization result on variable $A$ and $C$ in Figure \ref{fig:auto_graph_2}. Although the number of discretization edges is less than in Figure \ref{fig:auto_exp1_distr_1_3}, it still matches the raw data distribution very well.}
  \label{fig:auto_exp2_distr_1_3}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \scalebox{0.8}{
      \input{graph_wine_1}}
  \caption{Structure from prediscretized K2 structure learning to compare the methods for Wine dataset}
  \label{fig:wine_graph_1}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  %\begin{tabular}{cc}
  \scalebox{0.8}{
   \input{graph_housing_1}}
   %\end{tabular}
  \caption{Structure from prediscretized K2 structure learning to compare the methods for Housing dataset}
  \label{fig:housing_graph_1}
\end{figure}

\begin{table}
  \centering
  \caption{Discretization result of Housing dataset based on the graph in Figure~\ref{fig:housing_graph_1}. The first twelve rows show the numbers of intervals after discretization by the three method. The thirteenth row is the log-likelihood of the dataset based on the discretizatio results. The heuristic method is over sensitive and discretizes some variables into too many intervals. The MDL method is less sensitive and discretizes most variables by only one interval. The optimal Bayesian approach has a good control on number of intervals. \tim{Fig. vs. Figure}}
  \scalebox{1.0}{
  \input{table_housing_1.tex}
  }

@@

\begin{figure}[ht]
\centering
    % \begin{tabular}{cc}
      \input{graph_housing_exp1_distr_3_5}
    % \end{tabular}
  \caption{The discretization result on variable $C$ and $E$ of the network in Figure~\ref{fig:housing_graph_1}. The scatter points are jittered due to the repeated values in the dataset. Furthermore, these repeated values lead to high number of discretization edges, since repeated values can easily form discretized areas by their own.}
  \label{fig:housing_exp1_distr_3_5}
\end{figure}

\begin{figure}[ht]
\centering
     %\begin{tabular}{cc}
    \input{graph_housing_exp1_distr_8_5}
     %\end{tabular}
  \caption{The discretization result on variable $H$ and $E$ of the network in Figure~\ref{fig:housing_graph_1}. Data points are not over repeated on $H$, thus the number of edges on $H$ remains desirable.}
  \label{fig:housing_exp1_distr_8_5}
\end{figure}

@@

\begin{figure}[ht]
    % \begin{tabular}{cc}
    \centering
    \scalebox{0.8}{
      \input{graph_housing_2}}
    % \end{tabular}
  \caption{The learned discrete-valued Bayesian network from the Housing dataset. Variables $C$, $E$, $J$, $K$ have high numbers of intervals after discretization, since they are in a Markov blanket.
  \tim{Var with 1 interval?}
  }
  \label{fig:housing_graph_2}
\end{figure}

@@

\begin{figure}[ht]
\centering
     %\begin{tabular}{cc}
      \input{graph_housing_exp2_distr_3_5}
     %\end{tabular}
  \caption{The discretization result on variable $C$ and $E$ of the network in Figure~\ref{fig:housing_graph_2}. The high number of discretization edges is due to the repeated values on the dataset in variable $C$ and $E$.}
  \label{fig:housing_exp2_distr_3_5}
\end{figure}

@@

To quantatively discuss the sensitivity of the three methods, consider the Bayesian network in Figure \ref{fig:exp_discuss}, where $X$ is continuous-valued and $P_1$ and $P_2$ is discrete-valued. Let $\Pi_X = \{  P_1, P_2 \}$. Assume $D_X = \{1.0,2.0,3.0, \ldots N \}$. Therefore, the corresponding objective functions to discretize $X$ in the three methods are:
{\begin{small}
\begin{equation}
\begin{aligned}
\text{Obj}_{\text{MDL}} & = \overbrace{Z_1 \cdot k + \log(k) + \log {{N +k - 1}\choose{k -1}} }^{\text{Penalty Term}}+   \overbrace{\log(N) \cdot I(X^*, \Pi_X)}^{\text{Disc. Edge Positions Term}} \\
\text{Obj}_{\text{Bayesian}} &= \overbrace{Z_2 \cdot k + \log {{n_i + J_P - 1}\choose{J_P -1}}}^{\text{Penalty Term}} + \overbrace{\sum^k_{i=1} \log \left(   {{n_i !}\over{n^{(P)}_{i,1}! n^{(P)}_{i,2}! \cdots n^{(P)}_{i,J_P}!    }} \right) }^{\text{Disc. Edge Positions Term}}\\
\text{Obj}_{\text{Heuristic}} &= \overbrace{Z_2 \cdot k + \log {{n_i + J_{P_1} - 1}\choose{J_{P_1} -1}} + \log {{n_i + J_{P_2} - 1}\choose{J_{P_2} -1}} }^{\text{Penalty Term}}+\\
& \overbrace{\sum^k_{i=1}  \left[ \log \left(   {{n_i !}\over{n^{(P_1)}_{i,1}! n^{(P_1)}_{i,2}! \cdots n^{(P_1)}_{i,J_{P_1}}    }} \right)  + \log \left(   {{n_i !}\over{n^{(P_2)}_{i,1}! n^{(P_2)}_{i,2} \cdots n^{(P_2)}_{i,J_{P_2}} }} \right) \right]}^{\text{Disc. Edge Positions Term}}
\end{aligned}
\end{equation}
\end{small}}
, where $Z_1$ and $Z_2$ are two constants, $k$ is the number of intervals after discretization, $X^*$ is the discretized version of $X$, and $I(X^*, \Pi_X)$ is the mutual information between $X^*$ and its parent variables. Note that the third term in $\text{Obj}_{\text{MDL}}$ is approximated by $H(p)$ as in Equation \ref{eqn:MDL} in the original paper \citep{Friedman_1996}.

The sensitivities toward the value distribution in $D_{\Pi_X}$ are determined by the `` Disc. Edge Positions Term" in the three objective functions. Since the penaltis of the numbers of intervals after discretization in the three objective functions take the similar form, if the `` Disc. Edge Positions Term" varies more severely due to different positions of discretization edges, then the corresponding method is more sensitive. The `` Disc. Edge Positions Term" in $\text{Obj}_{\text{MDL}}$, which is made out of the mutual information, varies less severely than the terms in $\text{Obj}_{\text{Bayesian}}$ and $\text{Obj}_{\text{Heuristic}}$, since it senses the distribution in $D_{\Pi_X}$ by empirical probability distributions, instead of the factorial terms such in $\text{Obj}_{\text{Bayesian}}$ and $\text{Obj}_{\text{Heuristic}}$. Furthermore, the `` Disc. Edge Positions Term" in $\text{Obj}_{\text{Heuristic}}$ varies more aggressively than the term in $\text{Obj}_{\text{Bayesian}}$. Therefore, the MDL method is less sensitive and the heuristic is over sensitive toward the value distributions in other variables. The effect of sensitivity can be shown as the numbers of intervals after discretization in the previous experiment.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{acknowledgements}
%If you'd like to thank anyone, place your comments here
%and remove the percent signs.
%\end{acknowledgements}
